Supercomputer parallel computations in solving the applied optimization problems

### 2017-09-01 2017/ParOptBook/Chapter1.tex

Often, the constraints in the form of equations are also included in the non-linear programming problem statement. However, any equation $h(y)=0$ , first, could be represented formally as the system of two inequalities $h(y)\leq 0$  and $h(y)\geq 0$ . Second, in numerical solving an optimization problem on a computer the exact calculation of the equality $h(y)=0$ is often impossible, therefore, this equality is assumed to be computed  with some error $\delta>0$ , i.e., the inequality $\left|h(y)\right|\leq \delta$ is considered instead of the equation $h(y)=0$. Thus, taking into account the above reasons, one can claim that (\ref{eq:8})-(\ref{eq:10}) is the statement of  \textit {the general  problem of the nonlinear programming}.

If $m=0$ , i.e., there are no functional constraints, let us assume $Q=D$. The problem (\ref{eq:8})-(\ref{eq:10}) in this case will be called \textit {the unconstrained optimization problem}.

Let us consider an optimization problem  for the function
\begin{displaymath}
\label{eq:12}
\varphi(y_1,y_2)=-1.5y_1^2exp(1-y_1^2-20.25(y_1-y_2)^2)-
\end{displaymath}
\begin{displaymath}
\label{eq:12}
-(0.5(y_1-1)(y_2-1))^4exp(2-(0.5(y_1-1))^4-(y_2-1)^4)
\end{displaymath}
within the domain $0 \leq y_1\leq 4,\ \  -1\leq y_2\leq 3$, with the following constraints 
\begin{displaymath}
g_1(y_1,y_2)=0.01((y_1-2.2)^2+(y_2-1.2)^2-2.25)\leq 0,
\end{displaymath}
\begin{figure}[t]
\includegraphics[width=0.7\linewidth]{figures/figure1_1.png}
\caption{Example of a global optimization problem with disconnected admissible domain}
\label{1_fig_1}    
\end{figure}
\begin{displaymath}
g_1(y_1,y_2)=100(1-(y_1-2)^2/1.44-(0.5y_2)^2)\leq 0,
\end{displaymath}
\begin{displaymath}
g_1(y_1,y_2)=10(y_2-1.5-1.5sin(6.283(y_1-1.75)))\leq 0,
\end{displaymath}
as a particular example.
The set of the points satisfying the first constraint represents a circle with the boundary defined by the equation $g_1(y_1,y_2)=0$. The points satisfying the second constraint are located outside an ellipse defined by the equation $g_2(y_1,y_2)=0$. The points satisfying the third constraint are below the sinusoid $g_3(y_1,y_2)=0$. Thus, the feasible domain is disconnect and consists of three non-convex subdomains (the feasible domain is dark in Fig.\ref{1_fig_1}). The global minimum $\varphi(y_1^*,y_2^*)=-1.489$  is achieved at the point $(y_1^*,y_2^*)$  = (0,942, 0,944).

@@

The classical approach of the mathematical analysis suggests the following procedure for the analytical solution of the problem. Let us consider the one-dimensional case and assume that $\varphi(y)$ is a single argument function, which is piecewise smooth over the interval $[a,b]$. Then, the minimum of $\varphi(y)$ in $[a,b]$ can be achieved only at the points, where the derivative $\varphi'(y)=0$ , or where the derivative is discontinuous, or at the boundary points. Therefore, it is necessary to find all these points and to select from them the point with the lowest value. In other words, in order to solve the problem by this method, the following requirements should be satisfied:
\begin{enumerate}
\item{it is necessary to know the analytical formula for the function;}
\item{the function should be piecewise smooth;}
\item{the calculation of the derivative should be possible;}
\item{it is necessary to solve the equation$\varphi'(y)=0$;}
\item{the information on the derivative discontinuity points or a method for finding these ones are required.}
\end{enumerate}

@@

For this method
\begin{displaymath}
G_1(\Phi)=a, G_{k=1}(\Phi,\omega_k)=a+k\frac{b-a}{q}, k\geq 1,
\end{displaymath}
\begin{displaymath}
  H_k(\Phi,\omega_k) =
  \begin{cases}
    0, & k > q, \\
    1, & k \leq q,
  \end{cases}
\end{displaymath}
\begin{equation}
\label{eq:18}
e^k=\varphi_k^*,
\end{equation}
where
\begin{equation}
\label{eq:19}
\varphi_k^*=\min_{1\leq i\leq k} \varphi(y^i),
\end{equation}
or
\begin{displaymath}
e^k=(\varphi_k^*,y_k^*),
\end{displaymath}
where 
\begin{displaymath}
\label{eq:19}
y_k^*=arg\min_{1\leq i\leq k} \varphi(y^i).
\end{displaymath}

\subsection {Parallel optimization algorithms}
\label {subsec:1.2.2}
The computational scheme (\ref{eq:12})-(\ref{eq:17}) implies  the sequential execution of all the algorithm steps, that can be provided by the single processor computers. But if a multiprocessor computer is available, a parallelization of the computational scheme and, therefore, an acceleration of the optimization process can be realized. The parallelization of optimization methods could be done in several ways. It is possible to parallelize, in the first place, the calculation of the objective function, secondly, the implementation of the rules $G_k,E_k,H_k$ , and, finally, the scheme (\ref{eq:12})-(\ref{eq:17}) providing simultaneous execution of several trials. However, the parallelization of the objective function describing the optimized object is specific for different objective function models, and as a matter of fact, cannot be related to the general optimization scheme. The parallelization of the functionals  $G_k,E_k,H_k$ depends essentially on the class of algorithms. Besides, often these functionals are rather simple, and their parallelization is just of no use. Therefore, Therefore, the subject of the consideration in the present subsection  is the design of some \textit {fundamental principles} for a parallel execution of trials.  Let us use the term \textit {iteration} for the simultaneous (parallel) execution of several trials (each trial is executed by a separate processor). Let us denote the number of trials during the $n$-th iteration as $p(n)$ , and the total number of trials executed during all $n$ iterations as $k(n)$ . In other words, we assume that we have  $p$  processors at our disposal while executing the $n$-th iteration. Actually, the number of processors employed can be different for different iterations.

Let us consider the synchronous variant of the parallelization, where the transition to the next iteration is performed after full termination of the current one, i.e., after completing the last trial of the current iteration. In this case, it is obvious that $k(n)=p(1)+\ldots p(n),n\geq 1$.  Let us denote the vector of the trial coordinates at the $n$-th iteration as $u(n)$ , i.e.,
\begin{displaymath}
u(n)=(y^{k(n-1)+1},\ldots ,y^{k(n)}) .
\end{displaymath}
and the first $j$ coordinates of the vector $u(n)$  as $u_j(n)$ , i.e., 
\begin{displaymath}
u(n)=(y^{k(n-1)+1},\ldots ,y^{k(n-1)+j}),\ 1\leq j\leq p(n) .
\end{displaymath}

@@

Another approach based on the idea of dimensionality reduction, i.e., on building the schemes, where solving a multidimensional problem is reduced to solving one or several one-dimensional subproblems is more fruitful as compared with the reduction to the local problems. The following two efficient schemes of this kind are known:
\begin{enumerate}
\item{the nested optimization scheme  of optimization \cite{1_CarrHowe, 1_Evtushenko, 1_GerGriIsr, 1_GerGriGer, 1_GriIsrAIP, 1_GriIsrSerg, 1_GriIsrCEUR, 1_Piyavskij, 1_SergGriJCAA, 1_ShiOlaf, 1_StrMonRus, 1_StrSergMon2000, 1_vanDam},}
\item{the reduction of dimensionality based on the curves filling the space (Peano-type curves) \cite{1_Butz, 1_Goertzel, 1_HimOliPet, 1_LeraSergCNSNS, 1_LeraSergANM, 1_SergStrLeraMonogr, 1_StrMonRus,1_StrSergMon2000}.}
\end{enumerate}
The \textit{nested optimization scheme} of dimensionality reduction replaces solving the problem (\ref{eq:8})-(\ref{eq:10}) with solving a family of recursively nested one-dimensional subproblems, in which every evaluation of the objective function within a one-dimensional subproblem consists in solving a new one-dimensional subproblem of the next level of the recursion except the last one, where the value of the initial multidimensional objective function is calculated.

>>>>

Often, the constraints in the form of equations are also included in the non-linear programming problem statement. However, any equation $h(y)=0$ , first, could be represented formally as the system of two inequalities $h(y)\leq 0$  and $h(y)\geq 0$ . Second, in numerical solving an optimization problem on a computer the exact calculation of the equality $h(y)=0$ is often impossible, therefore, this equality is assumed to be computed  with some error $\delta>0$ , i.e., the inequality $\left|h(y)\right|\leq \delta$ is considered instead of the equation $h(y)=0$. Thus, taking into account the above reasons, one can claim that (\ref{eq:8})--(\ref{eq:10}) is the statement of  \textit {the general  problem of the nonlinear programming}.

If $m=0$ , i.e., there are no functional constraints, let us assume $Q=D$. The problem (\ref{eq:8})--(\ref{eq:10}) in this case will be called \textit {the unconstrained optimization problem}.

Let us consider an optimization problem  for the function
\begin{displaymath}
\label{eq:12}
\varphi(y_1,y_2)=-1.5y_1^2\exp(1-y_1^2-20.25(y_1-y_2)^2)-
\end{displaymath}
\begin{displaymath}
\label{eq:12}
-(0.5(y_1-1)(y_2-1))^4\exp(2-(0.5(y_1-1))^4-(y_2-1)^4)
\end{displaymath}
within the domain $0 \leq y_1\leq 4,\ \  -1\leq y_2\leq 3$, with the following constraints 
\begin{displaymath}
g_1(y_1,y_2)=0.01((y_1-2.2)^2+(y_2-1.2)^2-2.25)\leq 0,
\end{displaymath}
\begin{displaymath}
g_2(y_1,y_2)=100(1-(y_1-2)^2/1.44-(0.5y_2)^2)\leq 0,
\end{displaymath}
\begin{displaymath}
g_3(y_1,y_2)=10(y_2-1.5-1.5\sin(6.283(y_1-1.75)))\leq 0,
\end{displaymath}
as a particular example.

\begin{figure}[t]
\includegraphics[width=0.7\linewidth]{figures/figure1_1.png}
\caption{Example of a global optimization problem with disconnected admissible domain}
\label{1_fig_1}    
\end{figure}

The set of the points satisfying the first constraint represents a circle with the boundary defined by the equation $g_1(y_1,y_2)=0$. The points satisfying the second constraint are located outside an ellipse defined by the equation $g_2(y_1,y_2)=0$. The points satisfying the third constraint are below the sinusoid $g_3(y_1,y_2)=0$. Thus, the feasible domain is disconnect and consists of three non-convex subdomains (the feasible domain is dark in Fig.~\ref{1_fig_1}). The global minimum $\varphi(y_1^*,y_2^*)=-1.489$  is achieved at the point $(y_1^*,y_2^*)$  = (0,942, 0,944).

@@

The classical approach of the mathematical analysis suggests the following procedure for the analytical solution of the problem. Let us consider the one-dimensional case and assume that $\varphi(y)$ is a single argument function, which is piecewise smooth over the interval $[a,b]$. Then, the minimum of $\varphi(y)$ in $[a,b]$ can be achieved only at the points, where the derivative $\varphi'(y)=0$ , or where the derivative is discontinuous, or at the boundary points. Therefore, it is necessary to find all these points and to select from them the point with the lowest value. In other words, in order to solve the problem by this method, the following requirements should be satisfied:
\begin{enumerate}
\item{it is necessary to know the analytical formula for the function;}
\item{the function should be piecewise smooth;}
\item{the calculation of the derivative should be possible;}
\item{it is necessary to solve the equation $\varphi'(y)=0$;}
\item{the information on the derivative discontinuity points or a method for finding these ones are required.}
\end{enumerate}

@@

For this method
\begin{displaymath}
G_1(\Phi)=a, G_{k=1}(\Phi,\omega_k)=a+k\frac{b-a}{q}, k\geq 1,
\end{displaymath}
\begin{displaymath}
  H_k(\Phi,\omega_k) =
  \begin{cases}
    0, & k > q, \\
    1, & k \leq q,
  \end{cases}
\end{displaymath}
\begin{equation}
\label{eq:18}
e^k=\varphi_k^*,
\end{equation}
where
\begin{equation}
\label{eq:19}
\varphi_k^*=\min_{1\leq i\leq k} \varphi(y^i),
\end{equation}
or
\begin{displaymath}
e^k=(\varphi_k^*,y_k^*),
\end{displaymath}
where 
\begin{displaymath}
\label{eq:19}
y_k^*=\arg \min_{1\leq i\leq k} \varphi(y^i).
\end{displaymath}

\subsection {Parallel optimization algorithms}
\label {subsec:1.2.2}
The computational scheme (\ref{eq:12})--(\ref{eq:17}) implies  the sequential execution of all the algorithm steps, that can be provided by the single processor computers. But if a multiprocessor computer is available, a parallelization of the computational scheme and, therefore, an acceleration of the optimization process can be realized. The parallelization of optimization methods could be done in several ways. It is possible to parallelize, in the first place, the calculation of the objective function, secondly, the implementation of the rules $G_k,E_k,H_k$ , and, finally, the scheme (\ref{eq:12})--(\ref{eq:17}) providing simultaneous execution of several trials. However, the parallelization of the objective function describing the optimized object is specific for different objective function models, and as a matter of fact, cannot be related to the general optimization scheme. The parallelization of the functionals  $G_k,E_k,H_k$ depends essentially on the class of algorithms. Besides, often these functionals are rather simple, and their parallelization is just of no use. Therefore, Therefore, the subject of the consideration in the present subsection  is the design of some \textit {fundamental principles} for a parallel execution of trials.  Let us use the term \textit {iteration} for the simultaneous (parallel) execution of several trials (each trial is executed by a separate processor). Let us denote the number of trials during the $n$-th iteration as $p(n)$ , and the total number of trials executed during all $n$ iterations as $k(n)$ . In other words, we assume that we have  $p$  processors at our disposal while executing the $n$-th iteration. Actually, the number of processors employed can be different for different iterations.

Let us consider the synchronous variant of the parallelization, where the transition to the next iteration is performed after full termination of the current one, i.e., after completing the last trial of the current iteration. In this case, it is obvious that $k(n)=p(1)+\ldots + p(n), \; n\geq 1$.  Let us denote the vector of the trial coordinates at the $n$-th iteration as $u(n)$ , i.e.,
\begin{displaymath}
u(n)=(y^{k(n-1)+1},\ldots ,y^{k(n)}) .
\end{displaymath}
and the first $j$ coordinates of the vector $u(n)$  as $u_j(n)$ , i.e., 
\begin{displaymath}
u(n)=(y^{k(n-1)+1},\ldots ,y^{k(n-1)+j}),\ 1\leq j\leq p(n) .
\end{displaymath}

@@

Another approach based on the idea of dimensionality reduction, i.e., on building the schemes, where solving a multidimensional problem is reduced to solving one or several one-dimensional subproblems is more fruitful as compared with the reduction to the local problems. The following two efficient schemes of this kind are known:
\begin{enumerate}
\item{the nested optimization scheme  of optimization \cite{1_CarrHowe, 1_Evtushenko, 1_GerGriIsr, 1_GerGriGer, 1_GriIsrAIP, 1_GriIsrSerg, 1_GriIsrCEUR, 1_Piyavskij, 1_SergGriJCAA, 1_ShiOlaf, 1_StrMonRus, 1_StrSergMon2000, 1_vanDam},}
\item{the reduction of dimensionality based on the curves filling the space (Peano-type curves) \cite{1_Butz, 1_Goertzel, 1_HimOliPet, 1_LeraSergCNSNS, 1_LeraSergANM, 1_SergStrLeraMonogr, 1_StrMonRus,1_StrSergMon2000}.}
\end{enumerate}
The \textit{nested optimization scheme} of dimensionality reduction replaces solving the problem (\ref{eq:8})--(\ref{eq:10}) with solving a family of recursively nested one-dimensional subproblems, in which every evaluation of the objective function within a one-dimensional subproblem consists in solving a new one-dimensional subproblem of the next level of the recursion except the last one, where the value of the initial multidimensional objective function is calculated.

### 2017-09-01 2017/ParOptBook/Chapter8.tex

The regional economy (the district, region, or state, etc. level) was the object of investigations. For the mathematical description of the economical processes taking place at the regional level, a dynamic balance normative economic model has been proposed in Dorodnicyn Computing Centre of RAS [7]. This model is a general one. However, it includes a significant number of parameters  being specific to the different regions. It is possible to find these parameters (or \textit{to identify the model}) by solving the problem of minimization of the deviation of the calculated time series of the macroeconomic measures from the corresponding historic statistical data. The functions entering the problem statement are the nonlinear ones. As a result, the problem of identification of the mathematical economic model is the global optimization one.

@@

8. N.N. Olenev. Model of innovation potential of regional economy // Proceedings of international conference “Economy of depressed areas: problems and prospects of progress of regional economy”. Barnaul, Altai State University Press. 2007. pp.178188. (In Russian)

>>>>

The regional economy (the district, region, or state, etc. level) was the object of investigations. For the mathematical description of the economical processes taking place at the regional level, a dynamic balance normative economic model has been proposed in Dorodnicyn Computing Centre of RAS \cite{8_Olenev1999}. This model is a general one. However, it includes a significant number of parameters  being specific to the different regions. It is possible to find these parameters (or \textit{to identify the model}) by solving the problem of minimization of the deviation of the calculated time series of the macroeconomic measures from the corresponding historic statistical data. The functions entering the problem statement are the nonlinear ones. As a result, the problem of identification of the mathematical economic model is the global optimization one.

Once the model was identified, it could be applied to verify various possible scenarios of the regional economic development on the base of the experiments with the model. The conclusions and predictions obtained using the identified mathematical model, in principle, could be disposed experimentally. This enables to correct further the model in order to obtain more justified prognosis based on the model being improved continuously.

The problem of the identification of the multisector model of the regional economy presented here has been solved on the base of the data available for Nizhni Novgorod region, Russia. As it has been already mentioned above, the normative models include a large number of unknown parameters (the normatives of the distribution of the products and the normatives of distribution of the financial resources). In order to identify these parameters, it was proposed to apply the efficient parallel global optimization methods. The application of these algorithms on the supercomputer systems allows increasing the level of complexity of the mathematical models of the regional economy expressed by the number of the independent parameters. Carrying out the scenario computations based on the identified models of the regional economy allows more exact prognosis of the economic consequences of various strategic decisions.

\subsection{The description of the model of the regional economy}

The economy of Nizhni Novgorod region can be aggregated into three main sectors with approximately equal power:
\begin{itemize}
	\item the branches of the infrastructure, production and distribution of raw materials (agriculture, electric power industry, development, transportation, state administration, education, public health);
	\item the manufacturing branches (machine building and metal working, chemical and petrochemical industry, fuel industry, ferrous and nonferrous metallurgy, construction materials industry, timber, woodworking, and pulp-and-paper industry, consumer goods industry, food, pharmaceutical, chemical, microbiological industry, flour-and-cereals and feed mill industry, printing industry, scientific complex);
	\item the service branches (trading, operations with real estate, financial services, other services).
\end{itemize}

Let us select the following economic agents while designing the model: Regional Government, the producers represented by the three sectors selected above, the banking system, and the households of the region, the external vendors and consumers.

In spite of absence of rich natural resources in Nizhni Novgorod Region its economy is one of the most developed one with respect to the industry among the regions of Russian Federation. The processing branches of industry, a powerful military industrial complex, well-developed fundamental scientific complex are the base of the industry in Nizhni Novgorod Region and there are good prospects for the development of the high-tech industries and of the innovative branches.

The three-sector variant of the general balance model with supplies of products, factors of production, and financial resources in taxation and the illegal circulation of goods and finances \cite{8_Olenev2007} has been applied as the base for the design of the mathematical model of the regional economy. Remind that the producers in the model of Nizhni Novgorod Region economy are represented by three sectors: 
\begin{itemize}
	\item the infrastructure complex ($X$),
	\item the manufacturing branches complex ($Y$),
	\item the services, mortgage, financial, and trading branches complex ($Z$).
\end{itemize}
The producers (the sectors of the regional economy $X$, $Y$, $Z$) use labor, capital, and the intermediate production of the partner sectors. The producers deliver their products to the internal markets and to the external ones. The households $L$ offer the labor and consume the final products. The resellers $T$ redistribute the goods and financial flows. The banking system $B$ lends money to the producers with the purpose to receive the banking profit. Regional Government $G$ collects the taxes from the producers and the households and manages the budget expenditures. Each market is assumed to establish its own prices for each kind of products. The variations of the prices are assumed to be inversely proportional to the variation of the supplies of the corresponding products.

In order to account for the illegal circulation of the goods and money, the producers were assumed to divide the manufactured products into the legal part and the illegal one. The latter does not undergo the taxation. As a result, the producers receive two kinds of income: the ``legal'' and the ``illegal'' one. The ``illegal'' money can be laundered and the reserves undergo penalties.

All the consumers’ money is assumed to be legal. The consumers’ income is assumed to be distributed between the acquisition of the legal products and the illegal ones from all sectors according to the predefined percentages. The production sectors $m = X,\;Y,\;Z$ pay the profit tax $n_1$, the value added tax $n_2$, the excise tax on the total output $n_3^m$ , the unified social tax on the payroll $n_4$, and the exports customs duties $n_5$. The households $L$ within the framework of the model pay the import customs duties $n_6$ and the income tax on the wages $n_7$.

The criteria and the parameters of the model have the upper and lower indices. The upper indices are used for the agents and the lower ones for the valuables. The supplies of each valuable are distributed according to the normative  $a_i^{nm}$, which are the portions of the amount of the valuable $i$ being transferred from the agent $n$ to the agent $m$. The distribution of money is also performed according to some normative $b_i^{nm}$, which are the portions of the money resources of the agent $m$ transferred to the agent $n$ for the product $i$. The capital coefficients are also defined by the normatives $c_i^m$ , which are the normalized amount of the product $i$ spent for the producing a unit of a fund product for the agent $m$.

The identification of the parameters was performed by the comparison of the output time series of the model variables with the available statistical time series in 2000--2008. As a criterion of proximity of the calculated time series to the statistical ones, Theil index was applied, which has been calculated according to the formula
\[
U=\frac{\sqrt{\sum{(x_i-y_i)^2}}}{\sqrt{\sum{x_i^2}}+\sqrt{\sum{y_i^2}}}
\]
The close the index to zero, the close the series compared.

The detailed description of the designed model is beyond the scope of this paper. The additional information can be found in \cite{8_Gergel2011}. Here we note that the global optimization problem arising in the model identification process has $60$ independent parameters, $86$ nonlinear constraints, and a multiextremal objective function. As an illustration, a two-dimensional section of the objective function confirming its essential multiextremality is presented in Fig.~\ref{8_fig_7}.

\begin{figure}[t]
%\sidecaption[t]
\includegraphics[width=0.8\linewidth]{figures/8_7.png}
\caption{Two-dimensional section of the objective function}
\label{8_fig_7}     
\end{figure}

This problem has been solved on the cluster of Lobachevsky State University of Nizhni Novgorod using the parallel index method with the rotated evolvents (8 processors were employed). The estimate of optimum was obtained after $300 000$ iterations, the computation time of obtaining an estimate was $26$ minutes. Total computation time was $3$ hours.

Using the identified mathematical model of Nizhni Novgorod Region, the scenario computations reflecting the dynamics of the economic indicators of the region for various economic policies (the pessimistic scenario, the basic, and the optimistic one) have been carried out in Computing Centre of RAS.

\section*{A brief conclusion}

In this chapter, some applied global optimization problems have been considered. All of these problems related with a search for optimal parameters for nonlinear relationships, which describe phenomenon or process under consideration. The first example is the finding of the optimum profile of a vehicle wheel. Solution to this problem allows to increase the maximum possible speed of the vehicle. The second one is related with chemical research (finding the unknown kinetic parameters in the mathematical models of the complex chemical reactions). The last example belongs to the field of economics. The problem of identification of the parameters for the regional economic model has been solved. Optimal parameters found with the global search algorithm allow to carry out scenario forecast calculations for the regional economy.

@@

\bibitem{8_Olenev2007}
Olenev, N.N.: Model of innovation potential of regional economy. Proceedings of international conference ``Economy of depressed areas: problems and prospects of progress of regional economy''. Altai State University Press. 178--188 (2007) (In Russian)
